\chapter{INTRODUCTION}
\label{ch:introduction}

% Purpose (from template): tell the reader WHY this study exists. Start broad,
% narrow to the research question. Define terms on first use. Reference prior work.
%
% Target length: ~600-900 words for this chapter (project is 5,000-10,000 total).
% See report/STRUCTURE.md for the full per-section budget.

\section{Overview}
\label{sec:intro:overview}

Modern machine-learning systems represent text, images, and other media as
high-dimensional vectors called embeddings. Items with similar meaning are placed
near one another, so searching a collection becomes a nearest-vector problem.
This is vector search, the operation behind image search, recommendation, and
retrieval-augmented language models. When the query and stored items are
different media, such as text-to-image search, the task is cross-modal retrieval.

Quantum computers process information with qubits, whose states can interfere
with one another. This is sometimes described as trying many answers at once, but
that description is too simple: a measurement still returns one outcome, and the
algorithm must make wrong answers less likely and useful answers more likely.
Classical computers can also test many candidates in parallel when enough
processors are available. A useful quantum speed-up therefore depends on the
whole workflow, including data loading and measurement, not only on the quantum
circuit itself~\cite{hoefler2023practicality}.

Two quantum building blocks look relevant to vector search. Grover's algorithm
finds a marked item in an unstructured collection using far fewer queries than a
classical scan, and the swap test estimates how similar two quantum states are.
This study asks, through measurement rather than assertion, whether those
building blocks are of practical use for cross-modal vector search at the scale
and on the hardware available today. It is an investigation of where quantum
search does and does not fit, not an attempt to demonstrate a quantum advantage.

\section{Background of Study}
\label{sec:background}

% What is vector search, what is cross-modal retrieval, what is CLIP, and what
% does "quantum similarity" mean here. Two or three paragraphs. Cite the
% foundational papers (CLIP, Grover, Buhrman et al. swap test).

For cross-modal retrieval we use CLIP, a model that turns images and captions
into vectors in the same space~\cite{radford2021clip}. Once both media are
represented there, a query is answered by ranking image vectors by closeness to
the text vector. Exhaustive search compares the query with every stored vector
and costs $O(N)$ comparisons for $N$ items. FAISS implements this exact scan
efficiently~\cite{johnson2019faiss}, while approximate methods such as the HNSW
graph reach a near neighbour in roughly $O(\log N)$ graph
steps~\cite{malkov2020hnsw}.

The quantum side of the comparison rests on two ideas. Grover's algorithm
searches an unstructured collection of $N$ items for a marked one using only
about $\sqrt{N}$ queries to an oracle, a quadratic improvement over the linear
classical scan~\cite{grover1996}. In this context, an oracle is the part of the
algorithm that can recognise whether a candidate is the desired answer. The swap
test takes two quantum states and estimates the squared overlap between them from
a single measured qubit~\cite{buhrman2001swaptest}; because a normalised
embedding can be encoded in the amplitudes of a quantum state, that overlap is a
similarity score. Grover can therefore be read as a search method, and the swap
test as a similarity comparison.

Table~\ref{tab:early_cost} summarises the main algorithms before the detailed
methodology. The important point is that Grover's $O(\sqrt{N})$ search is better
than a classical exact scan, but not automatically better than HNSW, which is
already roughly logarithmic because it accepts an approximate nearest neighbour.

\begin{table}[htbp]
  \centering
  \caption[Algorithm cost summary]{High-level cost summary for the search
  methods used in the project. $N$ is the number of stored vectors and $M$ is the
  hybrid shortlist size. The quantum costs count the search or comparison step
  only; data loading is discussed below.}
  \label{tab:early_cost}
  \renewcommand{\arraystretch}{1.3}
  \begin{tabularx}{\linewidth}{l l >{\raggedright\arraybackslash}X}
    \toprule
    \textbf{Method} & \textbf{Cost} & \textbf{Meaning in this project} \\
    \midrule
    Brute-force cosine & $O(N)$ & Exact baseline; reads every stored vector. \\
    \addlinespace
    FAISS flat & $O(N)$ & Same exact scan implemented by FAISS. \\
    \addlinespace
    HNSW & $O(\log N)$ approx. & Fast approximate graph index; can miss the exact nearest neighbour. \\
    \addlinespace
    Swap test & $O(N)$ circuits & Quantum similarity estimate; a full ranking still compares against every candidate. \\
    \addlinespace
    Grover & $O(\sqrt{N})$ oracle calls & Faster than an exact scan on paper, but only after a useful oracle exists. \\
    \addlinespace
    Hybrid & $O(\log N + M)$ & HNSW searches; the swap test reranks a shortlist of $M$ candidates. \\
    \bottomrule
  \end{tabularx}
\end{table}

The difficulty is that these speed-ups assume the data is already available
inside the quantum computer. For Grover to search real vectors, an oracle would
have to recognise the closest item while the whole dataset is held in
superposition, which requires a qRAM that does not yet exist at usable scale;
loading the data one item at a time instead costs $O(N)$
and removes the very advantage the algorithm promises~\cite{aaronson2015readfine}.
This tension between an algorithm's asymptotic promise and the cost of getting
data into the machine is the central question the study examines.

\section{Problem Statement and Research Aims}
\label{sec:problem}

% Make the gap explicit: classical vector search is well-studied; the practical
% relevance of quantum vector search at NISQ scale is not. State what we set out
% to MEASURE - not what we set out to PROVE.

Classical vector search is mature and fast. Whether quantum methods add anything
at today's scale is less clear, especially because practical-advantage analyses
warn that input and output costs can remove the benefit of algorithms with only
quadratic speed-ups~\cite{hoefler2023practicality}. The gap this work addresses
is empirical: there are few like-for-like comparisons that run classical and
quantum similarity search on the same data while reporting both retrieval
accuracy and resource cost. We built such a comparison. A common interface lets
classical, quantum, and hybrid engines answer the same queries over the same
embeddings, so their rankings are scored identically.

We frame the work as a measurement rather than a proof of advantage. Present-day
devices are small and noisy, characteristic of current NISQ hardware, so we run
the quantum engines mostly on a classical simulator and validate the hardware
path with a single small run on a real quantum processor. The goal is to
determine, on a controlled cross-modal dataset, how the quantum and hybrid
engines compare with strong classical baselines, the established methods any new
approach must outperform to be worth adopting, and to state plainly the
conditions under which each approach is preferable.

\section{Research Questions and Objectives}
\label{sec:questions}

% List 3-4 measurable research questions. Each must map to something the
% benchmark harness already produces (MRR, oracle calls, circuit depth, etc.)
% so we never write a question we cannot answer with data we already have.

\begin{enumerate}[label=RQ\arabic*., leftmargin=*]
  \item \emph{Accuracy:} Does swap-test similarity reproduce the ranking of
        classical cosine similarity, and what is the shot budget required to do so?
  \item \emph{Scaling:} Does our Grover implementation reproduce the
        $\lfloor \pi\sqrt{N}/4 \rfloor$ oracle-call scaling in practice?
  \item \emph{Practicality:} Under what conditions, if any, would a
        quantum vector-search engine outperform the HNSW classical baseline on
        the same dataset?
  \item \emph{Fit:} Which part of cross-modal vector search, if any, is a good
        match for quantum computation once data loading and classical baselines
        are included?
\end{enumerate}

The research questions above state what we set out to measure; the objectives
below state what we set out to build in order to measure it. Each objective
therefore maps onto a concrete project output, and together they close the gap
identified in the problem statement, the absence of a like-for-like comparison
that runs classical and quantum similarity search on the same data while
reporting both accuracy and cost. We set four objectives:

\begin{enumerate}[leftmargin=*]
  \item To implement a suite of search engines spanning classical, quantum, and
        hybrid approaches behind a single common interface, so that one query can
        be answered by any of them and the results compared directly. This
        interface is what makes the comparison in RQ1 and RQ3 like-for-like.
  \item To build a reproducible benchmark harness and a controlled cross-modal
        dataset, and to collect a complete set of results across embedding
        dimensions and measurement shot counts, which supplies the measured
        accuracy and resource data that RQ1, RQ2, and RQ4 are answered from.
  \item To provide a web application that displays the engines' rankings side by
        side, so that the comparison central to the problem statement is open to
        inspection rather than reported only as aggregate numbers.
  \item To write a balanced analysis that answers each research question from the
        measured data, including the cases where quantum methods do not help, in
        keeping with the investigative rather than promotional aim of the study.
\end{enumerate}

\section{Summary}
\label{sec:intro:summary}

This chapter introduced cross-modal vector search and the two quantum primitives,
Grover's search and the swap test, whose practical relevance to that task this
study examines. It framed the work as an empirical measurement rather than an
attempt to demonstrate a quantum advantage, set out four research questions on
accuracy, scaling, practicality, and fit, and stated the four objectives that
answer them. The remainder of the report is organised as follows.
Chapter~\ref{ch:literature} reviews the relevant prior work on classical vector
search, cross-modal embeddings, and quantum similarity primitives.
Chapter~\ref{ch:methodology} describes the system, dataset, search engines, and
evaluation metrics, and explains how the benchmark was run.
Chapter~\ref{ch:results} reports the measured accuracy and resource results and
answers each research question in turn, and Chapter~\ref{ch:conclusion} draws the
study together, states its limitations, and points to future work.
