\chapter{LITERATURE REVIEW}
\label{ch:literature}

% Focused review. Do NOT write a history of quantum computing. The committee is
% not made of quantum experts (see project brief), so explain just enough for a
% CS-bachelor reader to follow the methodology chapter.
% Target length: ~800-1,200 words.

\section{Classical Vector Search}
\label{sec:lit:classical}

% Cover, in order of complexity: exact brute force, FAISS flat indices,
% approximate-nearest-neighbour graphs (HNSW). Frame HNSW as the production
% standard - it is the most important comparison point on slide 11 of the
% midterm deck.

The baseline for vector search is an exact scan: the query vector is compared
against every stored vector, and the closest are returned. This guarantees the
correct nearest neighbours but costs $O(N)$ comparisons per query, which becomes
the bottleneck as a collection grows. Facebook AI Similarity Search (FAISS) is a
widely used library that performs this exact search efficiently by exploiting the
vectorised instructions of modern processors, and extends it to very large
collections~\cite{johnson2019faiss}. When exact search is too slow, approximate
nearest-neighbour methods accept a small loss in recall for a large gain in
speed. The most widely used of these is the Hierarchical Navigable Small World
(HNSW) graph, which links each vector to a set of near neighbours across several
layers and answers a query by greedily descending the graph, reaching a near
neighbour in roughly $O(\log N)$ steps~\cite{malkov2020hnsw}. We use HNSW as the
strongest classical baseline in this report because it is the kind of fast,
approximate index a practical retrieval system would choose before considering a
quantum alternative. A quantum search that hopes to be useful must therefore
improve on a classical method that is already sublinear, not only on a simple
linear scan.

\section{Cross-Modal Embeddings}
\label{sec:lit:embeddings}

% One short subsection on CLIP. Explain contrastive learning at a high level
% and why text and image embeddings end up in the same metric space.

Cross-modal retrieval depends on embedding different kinds of media into a
common space. CLIP (Contrastive Language-Image Pre-training) does this for text
and images by training a text encoder and an image encoder together, so that an
image and its caption map to nearby vectors while mismatched pairs are pushed
apart~\cite{radford2021clip}. The variant we use pairs a Vision Transformer image
encoder (ViT-B/32) with a text encoder and produces 512-dimensional embeddings,
in which the cosine similarity between a text vector and an image vector measures
their relevance. This reduces text-to-image retrieval to ordinary
nearest-neighbour search in a single vector space. For the quantum engines, which
encode a vector into the amplitudes of a quantum state, we truncate these
512-dimensional embeddings to 64, 128, or 256 components to keep the number of
simulated qubits manageable, a trade-off examined in the methodology and results
chapters.

\section{Quantum Similarity Primitives}
\label{sec:lit:quantum}

% Swap test (Buhrman et al.), amplitude encoding state preparation, Grover.
% Be honest about what is "speedup" and what is "amortised state preparation
% cost" - the qRAM caveat goes here, not in the methodology.

Two quantum building blocks are relevant to similarity search. Grover's algorithm
searches an unstructured collection of $N$ items for one that satisfies a
condition checked by an oracle, and it does so with about $\lfloor (\pi/4)\sqrt{N}
\rfloor$ oracle queries, a quadratic improvement on the $N$ checks a classical
scan requires~\cite{grover1996}. The swap test, introduced in work on quantum
fingerprinting, compares two quantum states by placing an ancilla qubit in
superposition, conditionally swapping the two states, and measuring the ancilla;
the probability of reading the ancilla as zero is $(1 + |\langle\psi|\phi\rangle|^2)/2$,
so the test recovers the squared overlap of the states and with it a similarity
score~\cite{buhrman2001swaptest}.

Both techniques assume the data is already present as quantum states, and this
assumption carries the cost that the asymptotic figures hide. Encoding $N$
vectors into a quantum state, or building an oracle that recognises the closest
of them, requires reading all $N$ items. A quantum random-access memory has been
proposed to supply data items in superposition~\cite{giovannetti2008qram}, but no
such device exists at usable scale, and if the data must instead be loaded one
item at a time, the $O(N)$ preparation cost cancels Grover's quadratic advantage
at the level of the whole system~\cite{aaronson2015readfine}. Hoefler, H\"{a}ner,
and Troyer make the same practical point more broadly: quantum computers are
most promising for small-data problems with very large computation, while
big-data search is limited by input and output movement~\cite{hoefler2023practicality}.
The present work therefore treats Grover's scaling as a property of the search
step in isolation and is explicit about the data-loading cost it does not
include, a distinction returned to in the results chapter.

\section{Hybrid Classical--Quantum Architectures}
\label{sec:lit:hybrid}

% Brief discussion of recent hybrid retrieval pipelines that use a classical
% prefilter and a quantum reranker. Frame our hybrid_hnsw_swap_test engine as
% an instance of this pattern.

The retrieve-and-rerank cascade is a standard pattern in information retrieval: a
fast, inexpensive first stage selects a small set of candidates from the whole
collection, and a slower but more accurate second stage reorders only those
candidates. The cost of the expensive stage is then bounded by the size of the
shortlist rather than by the size of the collection. This pattern is a natural
fit for near-term quantum hardware. Quantum machine learning remains constrained
by considerable hardware and software challenges~\cite{biamonte2017qml}, so a
practical near-term design keeps the global, data-bound work on classical
hardware and reserves the quantum processor for a small, structured subproblem.

Our \texttt{hybrid\_hnsw\_swap\_test} engine is an instance of this cascade. The
classical HNSW index of Section~\ref{sec:lit:classical} selects a small candidate
pool in roughly $O(\log N)$ steps, and the swap test of
Section~\ref{sec:lit:quantum} reranks only those candidates by their estimated
similarity, executed on the Qiskit Aer simulator~\cite{qiskit2024}. Because the
swap test runs one circuit per candidate, confining it to the shortlist keeps the
number of circuit executions independent of the collection size, at $O(\log N +
M)$ for a shortlist of $M$ candidates. This places the quantum step where its
cost is bounded and where an estimation error can only reorder an already-relevant
shortlist rather than miss a relevant item entirely, an arrangement whose accuracy
and cost we evaluate in the results chapter.
