\chapter{RESEARCH METHODOLOGY}
\label{ch:methodology}

% Describe how the study was conducted, well enough for someone else to
% reproduce it. This chapter is the bulk of the report (~1,500-2,500 words).
% The "Use of AI Tools" subsection at the END is MANDATORY per handbook \S1.1.

\section{System Overview}
\label{sec:method:overview}

% One paragraph describing the modular monolith - data loader, embedding
% pipeline, engines, storage, API+UI. Reference Figure \ref{fig:architecture}.

The system is a modular monolith that performs cross-modal retrieval: given a
short text query such as ``a dog on a beach'', it returns the dataset images
whose content best matches that description. Text and images are never compared
directly. Instead, the same embedding model maps both into a shared vector
space, and retrieval becomes the problem of finding the dataset vectors closest
to the query vector. The defining property of the system is that the component
performing this closest-vector search is interchangeable. A classical engine, a
quantum engine, and a hybrid of the two all expose one common interface, so a
single query can be answered by any of them and the rankings placed side by
side. This makes the system an instrument for comparing search strategies
rather than a single retrieval product.

This interchangeability is enforced by a single design decision that runs
through the whole codebase: the strategy pattern. Each stage of the pipeline is
defined by an abstract interface, and concrete implementations plug into it.
Data loaders implement \texttt{BaseDataLoader}, embedding models implement
\texttt{EmbeddingGenerator}, and every search engine implements
\texttt{SearchEngineStrategy}, which requires just two operations,
\texttt{build\_index} and \texttt{search}. Because all engines satisfy the same
contract, the surrounding harness that loads data, generates embeddings, runs
queries, and records results is identical for every engine. Any difference we
measure between a classical and a quantum engine therefore reflects the engine
itself, not the code around it. Which engines run, at which vector dimensions,
and with how many measurement shots is selected in a configuration file
(\texttt{benchmarks.yaml}) rather than by editing source, so an experiment is
defined by configuration rather than by code changes.

The pipeline is organised into five phases, shown in
Figure~\ref{fig:architecture}. In the first phase, data loading, a
\texttt{DirectoryDataLoader} reads the image files and the accompanying
ground-truth pairs that define which image each query should retrieve. In the
second phase, the embedding pipeline, a CLIP model converts every image and
every text query into a vector. In the third phase the search engines build
their indexes over these vectors and answer queries, each engine applying its
own strategy, whether classical, quantum, or hybrid. In the fourth phase the
outcome of every run, including its accuracy and resource measurements, is
written to a PostgreSQL database. The fifth phase exposes the system through a
FastAPI backend and a React interface, which lets a user issue a query and
inspect the rankings returned by each engine. The sections that follow describe
each phase in turn: the dataset and ground truth in
Section~\ref{sec:method:dataset}, the embedding pipeline in
Section~\ref{sec:method:embeddings}, and the individual search engines in
Section~\ref{sec:method:engines}.

\begin{figure}[ht]
  \centering
  \begin{tikzpicture}[
      font=\small,
      box/.style={draw, rounded corners, align=center, inner sep=5pt,
        text width=58mm, fill=gray!8},
      eng/.style={draw, rounded corners, align=center, inner sep=4pt,
        text width=30mm, minimum height=12mm, font=\footnotesize},
      classical/.style={eng, fill=blue!10, draw=blue!55},
      quantum/.style={eng, fill=violet!12, draw=violet!55},
      hybrid/.style={eng, fill=green!12, draw=green!50!black!70},
      arr/.style={-{Latex[length=2.2mm]}, semithick}
    ]
    \node[box] (load) {\textit{Phase 1: Data loading}\\[2pt]
      \textbf{DirectoryDataLoader}\\ images and ground-truth pairs};
    \node[box, below=7mm of load] (embed) {\textit{Phase 2: Embedding}\\[2pt]
      \textbf{CLIPEmbeddingModel}\\ ViT-B/32, one vector per image and query};
    \node[box, below=7mm of embed] (strategy) {\textit{Phase 3: Search engines}\\[2pt]
      \textbf{SearchEngineStrategy}\\ \texttt{build\_index} / \texttt{search}};

    \node[quantum, below=11mm of strategy] (qu) {\textbf{Quantum}\\ swap test, Grover};
    \node[classical, left=5mm of qu] (cl) {\textbf{Classical}\\ brute force, FAISS flat, FAISS HNSW};
    \node[hybrid, right=5mm of qu] (hy) {\textbf{Hybrid}\\ HNSW + swap test};

    \node[box, below=13mm of qu] (db) {\textit{Phase 4: Storage}\\[2pt]
      \textbf{PostgreSQL + pgvector}\\ benchmark results};
    \node[box, below=7mm of db] (api) {\textit{Phase 5: API and interface}\\[2pt]
      \textbf{FastAPI + React}\\ side-by-side rankings per engine};

    \draw[arr] (load) -- (embed);
    \draw[arr] (embed) -- (strategy);
    \draw[arr] (strategy.south) -- (cl.north);
    \draw[arr] (strategy.south) -- (qu.north);
    \draw[arr] (strategy.south) -- (hy.north);
    \draw[arr] (cl.south) -- (db.north);
    \draw[arr] (qu.south) -- (db.north);
    \draw[arr] (hy.south) -- (db.north);
    \draw[arr] (db) -- (api);
  \end{tikzpicture}
  \caption{End-to-end pipeline. Data loading and CLIP embedding feed a common
  engine-strategy interface, behind which the classical, quantum, and hybrid
  engines are interchangeable. Every run is recorded in PostgreSQL and surfaced
  through the FastAPI and React interface.}
  \label{fig:architecture}
\end{figure}

\section{Methodology Diagram}
\label{sec:method:diagram}

% Per project brief: "Trebamo diagram metodologije napravit, kao diagram how we
% did all od this step by step." Insert it here.

\begin{figure}[ht]
  \centering
  \begin{tikzpicture}[
      font=\small,
      procbox/.style={draw, rounded corners, align=center, inner sep=4pt,
        text width=56mm, fill=gray!8, minimum height=11mm},
      store/.style={draw, align=center, inner sep=4pt, text width=56mm,
        fill=blue!8, draw=blue!50, minimum height=10mm},
      ibm/.style={draw, rounded corners, align=center, inner sep=4pt,
        text width=38mm, fill=orange!18, draw=orange!75!black,
        minimum height=16mm, font=\footnotesize},
      lane/.style={font=\footnotesize\itshape, text=gray, rotate=90},
      arr/.style={-{Latex[length=2.2mm]}, semithick}
    ]
    \node[procbox] (imp) {\textbf{import\_dataset.py}\\ Flickr30k subset + ground truth};
    \node[procbox, below=6mm of imp] (idx) {\textbf{index\_dataset.py}\\ CLIP vectors stored in PostgreSQL};
    \node[procbox, below=10mm of idx] (run) {\textbf{run\_benchmarks.py}\\ engine $\times$ dim $\times$ shots on AerSimulator};
    \node[store, below=14mm of run] (res) {\textbf{benchmark\_results} (PostgreSQL)\\ one row per run: MRR, oracle calls, depth, qubits};
    \node[procbox, below=10mm of res] (rep) {\textbf{generate\_report.py}\\ benchmark\_report.md, the source of truth};

    \node[ibm, right=9mm of run] (ibm) {\textbf{run\_ibm\_validation.py}\\ hybrid swap test on a real IBM QPU\\ (dim 2, 2 candidates, 32 shots)};

    \node[lane] at ([xshift=-7mm]$(imp.west)!0.5!(idx.west)$) {Data prep};
    \node[lane] at ([xshift=-7mm]$(run.west)!0.5!(res.west)$) {Execution};
    \node[lane] at ([xshift=-7mm]rep.west) {Reporting};

    \draw[arr] (imp) -- (idx);
    \draw[arr] (idx) -- (run);
    \draw[arr] (run) -- (res);
    \draw[arr] (res) -- (rep);
    \draw[arr] (ibm.south) |- (res.east);
  \end{tikzpicture}
  \caption{Methodology flow. Four scripts run in order, from dataset import
  through CLIP indexing, the simulator benchmark sweep, and report generation.
  The IBM hardware validation is a one-off side arm that writes its single
  result into the same database table.}
  \label{fig:methodology}
\end{figure}

The study follows a fixed sequence of steps, shown in
Figure~\ref{fig:methodology}, each implemented by a single script so the whole
experiment can be reproduced from a clean checkout. The first two steps prepare
the data. The script \texttt{import\_dataset.py} downloads a fixed subset of the
Flickr30k dataset and, in the same pass, writes the ground-truth file that
records which image each text query should retrieve. This download is
deterministic: every run regenerates the same image set and the same
query-to-image pairing, so the results do not drift between machines. The second
script, \texttt{index\_dataset.py}, loads each image, passes it through the CLIP
model to obtain a vector, and stores that vector in PostgreSQL. After these two
steps the database holds one embedding per image, ready to be searched.

With the embeddings in place, \texttt{run\_benchmarks.py} carries out the main
experiment. It iterates over every combination of engine, vector dimension, and
shot count listed in \texttt{benchmarks.yaml}, and for each combination it
answers all of the ground-truth queries. The quantum and hybrid engines execute
their circuits on Qiskit's \texttt{AerSimulator}, a noiseless simulator that
isolates the effect of finite sampling (shot noise) from the physical noise of
real hardware. Every individual run is recorded as one row in the
\texttt{benchmark\_results} table, together with its accuracy and its resource
cost: the Mean Reciprocal Rank of the retrieved images, and, for the quantum
engines, the number of oracle calls, the circuit depth, and the qubit count.
Because the configuration file drives the sweep, widening the experiment to a
new dimension or shot count is a change to that file alone, not to the code.

Alongside the simulator sweep, \texttt{run\_ibm\_validation.py} runs the hybrid
swap-test engine once on real IBM quantum hardware, at the smallest workable
size: two dimensions, two candidate images, and 32 shots. Its purpose is
narrow, to confirm that the swap-test circuit executes on a physical device and
returns a sensible result, rather than to measure scaling. The single outcome
is written to the same \texttt{benchmark\_results} table, tagged as the
\texttt{hybrid\_hnsw\_swap\_test\_ibm} engine, so it sits beside the simulator
runs without being confused for one.

The final step turns the stored runs back into something readable. The script
\texttt{generate\_report.py} queries the \texttt{benchmark\_results} table and
produces a Markdown summary, \texttt{benchmark\_report.md}, that groups the runs
by engine and reports accuracy per dimension and per shot count. This report is
the single source of truth for the Results chapter: every figure quoted in
Chapter~\ref{ch:results} traces back to a row this script read from the
database, rather than to any number recorded by hand. Keeping the analysis in
one generated file, derived directly from the stored runs, is what lets the
whole study be regenerated by re-running the four scripts in order.

\section{Dataset and Ground Truth}
\label{sec:method:dataset}

% Flickr30k subset, how it is loaded, what counts as a "correct" hit. Be
% explicit that ground truth is image-caption pairs from the dataset itself,
% not human relevance judgements.

[DATASET \,--\, source, size, how queries were derived, what counts as a
correct retrieval. Reference backend/data/ground\_truth.jsonc.]

\section{Embedding Pipeline}
\label{sec:method:embeddings}

% CLIP ViT-B/32, L2-normalisation, dimension truncation (64 / 128). Explain why
% truncation is necessary for the swap test on a CPU simulator.

[EMBEDDINGS \,--\, model, normalisation, truncation rationale.]

\section{Search Engines}
\label{sec:method:engines}

% One subsubsection per engine; keep each to a single paragraph plus formula(s)
% where relevant. Cross-reference docs/ENGINES_GUIDE.md as the authoritative
% source on parameter ranges.

\subsection{Classical Baselines}
\subsubsection{Brute-Force Cosine}
[BRUTE FORCE \,--\, NumPy dot product against every vector. Ground truth.]

\subsubsection{FAISS Flat (Exact)}
[FAISS FLAT \,--\, SIMD-vectorised exact search.]

\subsubsection{FAISS HNSW (Approximate)}
[FAISS HNSW \,--\, multi-layer proximity graph, $O(\log N)$ retrieval.]

\subsection{Quantum Engines}
\subsubsection{Qiskit Swap Test}
% Include the formula here. Equations must be numbered per FENS template.
The swap test estimates the squared inner product of two amplitude-encoded
vectors $\ket{\psi}$ and $\ket{\phi}$ via a single ancilla measurement:
\begin{equation}
P(\text{ancilla}=0) = \frac{1 + |\langle\psi|\phi\rangle|^2}{2}.
\label{eq:swaptest}
\end{equation}

[SWAP TEST \,--\, explain circuit (Hadamard, controlled-SWAP, Hadamard,
measure), shot-noise scaling, and the choice of AerSimulator over IBM
hardware.]

\subsubsection{Qiskit Grover (Hardcoded Oracle)}
[GROVER \,--\, $\lfloor\pi\sqrt{N}/4\rfloor$ oracle calls, how we measure that
empirically, why the oracle is classically pre-specified (qRAM caveat).]

\subsubsection{Qiskit Grover with Quantum State Preparation}
[GROVER (quantum prep) \,--\, contrast with the hardcoded variant.]

\subsubsection{Hybrid HNSW + Swap Test}
[HYBRID \,--\, HNSW prefilter, swap-test rerank. Discuss the rationale.]

\subsubsection{IBM Hardware Validation}
[IBM HARDWARE \,--\, the dimension-2, 2-candidate, 32-shot run on real QPU.
Frame as a sanity check, not a scaling study.]

\section{Evaluation Metrics}
\label{sec:method:metrics}

% Define MRR formally, then define oracle calls. Explain why wall-clock is per-
% engine only - the simulator-time issue.

[MRR DEFINITION \,--\, formula + intuition.]

\begin{equation}
\text{MRR} = \frac{1}{|Q|} \sum_{q \in Q} \frac{1}{\text{rank}_q},
\label{eq:mrr}
\end{equation}
where $\text{rank}_q$ is the position of the first relevant result for query
$q \in Q$ and $1/\text{rank}_q \coloneqq 0$ if no relevant result is in the top-$k$.

[ORACLE CALLS, CIRCUIT DEPTH, QUBIT COUNT \,--\, define each, state which
engines populate which fields.]

\section{Experimental Configuration}
\label{sec:method:config}

% Table summarising the cartesian product of (engine, dimension, shots, layers)
% actually run. Pull values from backend/config/benchmarks.yaml.

\begin{table}[ht]
  \centering
  \caption{Benchmark configuration grid.}
  \label{tab:config}
  \begin{tabular}{ll}
    \toprule
    \textbf{Parameter} & \textbf{Values} \\
    \midrule
    Engines             & [PULL FROM benchmarks.yaml] \\
    Dimensions          & 64, 128 \\
    Shots               & 512, 1024 \\
    Layers              & 1 \\
    Top-$k$             & 10 \\
    \bottomrule
  \end{tabular}
\end{table}

\section*{Use of AI Tools}
\label{sec:method:ai}
\addcontentsline{toc}{section}{Use of AI Tools}

% MANDATORY per FENS handbook \S1.1. Be specific - generic disclaimers fail
% the originality check. State which tools were used for which tasks.

[USE OF AI TOOLS \,--\, list every AI-assisted activity: code generation /
review, prose drafting, diagram drafting, brainstorming. State explicitly which
parts of the algorithm design, implementation, and analysis are the authors'
own. Responsibility for correctness rests with the authors. See
GUIDELINES.md \S AI Disclosure for the exact phrasing template.]
