\chapter[SUSTAINABILITY, INCLUSIVITY, AND SOCIETAL IMPACT]{SUSTAINABILITY, INCLUSIVITY, AND\\SOCIETAL IMPACT}
\label{ch:sustainability}

% Short reflection (one or two paragraphs is enough per template). SDG mapping
% optional. The poster grading form awards up to +10% bonus for this, so worth
% getting right.

This project was designed to be modest in its use of resources. The quantum
engines ran on a CPU-based circuit simulator over a deliberately small dataset of
twenty images, rather than on large-scale GPU training or extended
quantum-hardware time, so the entire benchmark sweep completes on a single
workstation; the one run on real quantum hardware consumed eighty seconds of a
free-tier quota. The data raises no privacy concerns, as Flickr30k is a public
research dataset of captioned images and contains no personal identifying
information that we collect or process. The work is openly reproducible: the
source code, the benchmark harness, and the dataset configuration are publicly
available, and the harness is deterministic, so the reported results can be
regenerated from scratch.

Beyond these practical points, the study makes a small contribution to how
quantum computing is discussed. Practical-advantage analyses warn that quantum
speed-ups are limited by data movement and by the size of the speed-up itself
\cite{hoefler2023practicality}; this project shows that issue in a concrete
retrieval system. By measuring where quantum search helps and where it does not,
the work offers a careful explanation rather than a general claim that quantum
methods are always useful. In the language of the United Nations Sustainable
Development Goals, the project speaks most directly to SDG~4 (quality education),
as an openly available teaching resource for quantum and classical retrieval, and
to SDG~9 (industry, innovation, and infrastructure), as an evidence-based
assessment of a candidate technology's readiness.
