\chapter{CONCLUSION}
\label{ch:conclusion}

% Summarise findings, highlight contributions, point to future work.
% Target length: ~300-500 words. No new results may appear here.
% Structure per the supervisor's review: 6.1 Summary, 6.2 Limitations (moved
% here from the Results chapter), 6.3 Future Work. No Overview subsection.

\section{Summary}
\label{sec:conclusion:summary}

This study built a single framework in which classical, quantum, and hybrid
engines answer the same cross-modal queries over the same embeddings, and used it
to measure where quantum similarity search helps. In the simulator benchmark,
retrieval accuracy mainly depended on the embedding dimension rather than on
whether an engine was quantum: the swap test reproduced classical accuracy once
given a sufficient measurement budget (RQ1), and our Grover implementation
reproduced its theoretical $\lfloor (\pi/4)\sqrt{N} \rfloor$ oracle-call count
exactly at the size we tested (RQ2). No engine that performed the search on a
quantum circuit outperformed the classical HNSW baseline on this dataset; the
only quantum-containing engine to match it, the hybrid, did so by delegating the
search to the classical index (RQ3). The reason is that vector retrieval is
dominated by reading the dataset, while quantum advantage is most plausible when
a small amount of data requires very large computation. Under the methods we
tested, cross-modal search is therefore not a strong target for a quantum search
engine, although a quantum reranker can still be studied as a bounded second
stage (RQ4).

This work delivers four things. The first is a reproducible benchmarking framework
that runs seven engines, classical, quantum, and hybrid, behind one interface and
records every result in a database. The second is an empirical verification of
textbook quantum behaviour on a real retrieval task: the swap test's convergence
to the classical ranking, and Grover's oracle-call scaling. The third is a web
application that displays the engines' rankings side by side, making the
comparison inspectable rather than abstract. The fourth, and the one we consider
most important, is the honest framing itself: stating plainly where quantum search
does not help, and why, is a contribution in a setting where that question is
often left unasked.

\section{Limitations}
\label{sec:conclusion:limitations}

Several constraints bound the results. The dataset has twenty images and one
query per image, which kept the sweep simulable and the target unambiguous but
limited statistical power. HNSW did not have enough vectors to exercise its
approximation, so its match with exact search holds only here. Except for the
IBM sanity check, the quantum engines ran on a classical simulator, so the
results do not measure present-day hardware noise or latency. Embeddings were
truncated from 512 to at most 256 dimensions to fit the qubit budget, so absolute
MRR values are comparisons under a shared handicap. Grover used a classically
precomputed oracle, and its oracle-call scaling was checked at one dataset size,
not across a range.

\section{Future Work}
\label{sec:conclusion:futurework}

Several directions would extend the study. A larger dataset, drawn from the full
Flickr30k set or an ImageNet subset, would exercise the HNSW approximation that
twenty vectors leave untested and give the accuracy figures more statistical
weight. A wider dimension sweep, and a study of variational state-preparation
circuits in place of the hardcoded oracle, would probe the quantum engines closer
to their intended use. Comparing against a production vector store such as
pgvector's HNSW would add a third classical baseline, and a characterisation of
noise on real quantum hardware, beyond the single validation run reported here,
would show how the simulated behaviour degrades on present-day devices.
